---
title: "Cyclistic"
author: "SAMPSON OKYERE"
date: "12/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cyclistic

Install needed packages
```{r}
  install.packages("tidyverse")
  install.packages("lubridate")
  library(tidyverse)
  library(lubridate)
```

Import data
```{r}
  setwd("~/Data_Cyclistic_Case_Study_1")
  trip_data_2020_12 <- read.csv("202012-divvy-tripdata.csv")
  trip_data_2021_01 <- read.csv("202101-divvy-tripdata.csv")
  trip_data_2021_02 <- read.csv("202102-divvy-tripdata.csv")
  trip_data_2021_03 <- read.csv("202103-divvy-tripdata.csv")
  trip_data_2021_04 <- read.csv("202104-divvy-tripdata.csv")
  trip_data_2021_05 <- read.csv("202105-divvy-tripdata.csv")
  trip_data_2021_06 <- read.csv("202106-divvy-tripdata.csv")
  trip_data_2021_07 <- read.csv("202107-divvy-tripdata.csv")
  trip_data_2021_08 <- read.csv("202108-divvy-tripdata.csv")
  trip_data_2021_09 <- read.csv("202109-divvy-tripdata.csv")
  trip_data_2021_10 <- read.csv("202110-divvy-tripdata.csv")
  trip_data_2021_11 <- read.csv("202111-divvy-tripdata.csv")
```
Check column names to ensure we can join all the data
```{r}
  colnames(trip_data_2020_12)
  colnames(trip_data_2021_01)
  colnames(trip_data_2021_02)
  colnames(trip_data_2021_03)
  colnames(trip_data_2021_04)
  colnames(trip_data_2021_05)
  colnames(trip_data_2021_06)
  colnames(trip_data_2021_07)
  colnames(trip_data_2021_08)
  colnames(trip_data_2021_09)
  colnames(trip_data_2021_10)
  colnames(trip_data_2021_11)
```
Combine data into one set.
```{r}
all_trips <- rbind(trip_data_2020_12, trip_data_2021_01, trip_data_2021_02, trip_data_2021_03, trip_data_2021_04, trip_data_2021_05, trip_data_2021_06, trip_data_2021_07, trip_data_2021_08, trip_data_2021_09, trip_data_2021_10, trip_data_2021_11)
```

Inspect data to find cleaning opportunities
```{r}
colnames(all_trips)
```
```{r}
nrow(all_trips)
```
```{r}
dim(all_trips)
```
```{r}
head(all_trips)
```
```{r}
tail(all_trips)
```
```{r}
str(all_trips)
```
```{r}
summary(all_trips)
```
```{r}
table(all_trips$start_station_name)
```
```{r}
table(all_trips$rideable_type)
```

Note stations "DIVVY CASSETTE REPAIR MOBILE STATION" and "HUBBARD ST BIKE CHECKING (LBS-WH-TEST)"

Add columns for date, month, day and year of each ride
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```
Add ride_length calculation in seconds and convert to number
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
```
Check for anomalies in ride_length data
```{r}
summary(all_trips$ride_length)
```

Remove bad data and put into v2 dataframe (less than 60 seconds, test or repair station, docked bikes)
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "DIVVY CASSETTE REPAIR MOBILE STATION" | all_trips$start_station_name == "HUBBARD ST BIKE CHECKING (LBS-WH-TEST)" | all_trips$rideable_type == "docked_bike" | all_trips$ride_length < 60),]
```
Double check for anomalies
```{r}
summary(all_trips_v2$ride_length)
```
Compare members vs casuals
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```
Correct order for day of the week
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Average ride time for members vs casuals
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
Rides by member vs casual and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  group_by(member_casual, weekday) %>%
  summarise(number_of_rides = n()	 
  ,average_duration = mean(ride_length)) %>%
  arrange(member_casual, weekday)
```
Show number of rides by rider type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```
Show average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```
Find only rides longer than one day per rider type
```{r}
long_trips <- all_trips_v2[(all_trips_v2$ride_length >= 86400),]
```

Show trips more than one day per rider type
```{r}
long_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```


Export summary files for further analysis
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = '~/Data_Cyclistic_Case_Study_1/avg_ride_length.csv')
```
```{r}
long_counts <- aggregate(long_trips$ride_length ~ long_trips$member_casual + long_trips$day_of_week, FUN = length)
write.csv(long_counts, file = '~/Data_Cyclistic_Case_Study_1/long_ride_length.csv')
```
```{r}
ride_counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = length)
write.csv(ride_counts, file = '~/Data_Cyclistic_Case_Study_1/number_of_rides.csv')
```


